% =============================================================================
% CHAPTER 5: Conclusions and Future Work
% =============================================================================
% SPDX-FileCopyrightText: 2026 Frank Langbein <frank@langbein.org>
% SPDX-License-Identifier: LPPL-1.3c
%
% Suggested sections: Conclusions; Future work. Do not introduce new material in
% Conclusions---summarise aims and main results. Future work: next steps, open
% questions, what you would do with more time. Optional: short summary of
% contributions (as part of Conclusions or a bullet list).
% =============================================================================

% This chapter concludes the dissertation. Summarise the aims, main results, and what was learned; then outline future work and improvements.

\section{Conclusions}\label{sec:conclusions}
% Summary of aims, key results, and main takeaways. Optional: list of contributions.
The project's main aim was to establish the most computationally efficient \gls{ilp} model for solving the NP-Complete Battleship Solitaire problem. This aim was successfully met, accomplishing the formulation, implementation, and evaluation of two distinctly different mathematical models: a dense-matrix Cell model and a sparse-matrix Ship model.

All pre-defined secondary objectives were also fully met in the course of this work. The project delivered a solid software architecture, which includes a custom \gls{csplib} parser, a procedural puzzle generator capable of producing instances up to $30 \times 30$ in size, and a thread-safe decoupled \gls{gui} used to present and interact with the solver's outputs.

The evaluation produced definitive, scale-tested results. The pre-processed Edge-Based Set Packing formulation of the Ship-based model demonstrated overwhelming computational dominance. On the largest \num{900}-cell instances existing in the generated dataset, it solved every puzzle in under \SI{2}{\second}, with an average of approximately \SI{0.4}{\second}. In contrast, the Cell-based model based on Big-M shape transition logic suffered severe exponential degradation, requiring a maximum of over \SI{23}{\second} for the same configurations.

Most importantly, the evaluation isolated the reasoning behind this degradation. By tracking the branch-and-bound node counts, the research proved that the Cell model's slowness was not caused by deep search-tree branching, but rather by extreme matrix bloat stalling Gurobi's root-node \gls{lp} relaxations. Furthermore, the constraint ablation study successfully quantified the `Overhead vs. Pruning' paradox, demonstrating that the polyhedral tightening via valid inequalities is only profitable at macroscopic scales (in this study, between 625 and 900 cells). 

Ultimately, this project demonstrates that reframing the mathematical paradigm of a spatial logic puzzle yields vastly superior computational returns compared to micro-optimising a flawed baseline.

\section{Future work}\label{sec:future}
% Next steps, open questions, and what you would do with more time or resources.

While this project successfully mapped the efficiency of these \gls{ilp} formulations up to $30 \times 30$ grids, the findings provide a clear starting point for future computational research into Discrete Tomography.

\subsection{Cross-Solver Benchmarking}\label{subsec:cross-solver-benchmarking}
As identified in the project limitations, the mathematical complexity metrics observed were inherently tied to Gurobi's presolve heuristics. Future work should implement a cross-solver benchmarking suite integrating open-source optimisation engines, such as CBC~\citep{coin-orcbc} or SCIP~\citep{achterberg2009scip}.

Evaluating the formulations across different solver backends would determine whether the Ship model's root-node efficiency is a universal mathematical property of Set Packing, or a byproduct of Gurobi's specific algorithmic bias.

\subsection{Statistical Resilience through Repeated Trials}\label{subsec:statistical-resilience}
A practical limitation acknowledged in Section~\ref{subsec:valid-inequalities-impact} is that each puzzle was solved exactly once per solver variant, so the recorded solve times include unquantified run-to-run variance from background processes, cache state, and Gurobi's internal randomisation. On the $10 \times 10$ ablation table this variance is large enough relative to the inter-configuration differences (1--4\,ms) that the table cannot confidently rank individual constraints.

A natural extension is to run each (instance, solver) configuration $k$ times, with $k$ chosen so that the standard error of the mean falls comfortably below the smallest effect of interest. With $k = 10$ repetitions per configuration the constraint ablation table would gain mean and standard-deviation columns that allow a Welch's $t$-test to be applied between rows. The $\sim$\SI{1.3}{\second} saving observed at \num{900} cells in Figure~\ref{fig:cuts_comparison} would be tested for significance against the same noise floor that masks the smaller effects.

A first repetition would be discarded as a warm-up run, isolating Gurobi's lazy-initialisation cost and the operating system's first-touch caching from the steady-state measurement that Chapter~\ref{ch4} reports. This refinement does not require changes to either ILP formulation, only to the evaluator's outer loop, and is therefore an inexpensive way to increase the statistical strength of the conclusions before any larger structural changes (such as the cross-solver work in Section~\ref{subsec:cross-solver-benchmarking}) are undertaken.

\subsection{A Hybrid ILP Formulation}\label{subsec:hybrid-ilp-formulation}
The most significant trade-off encountered in this project was time versus space. The Ship model provides rapid solve times, at the cost of exponential memory consumption during the candidate generation process.

In future work, a Hybrid \gls{ilp} model could be engineered to mitigate this memory ceiling on theoretical grids larger than $50 \times 50$. This approach would use Set Packing to efficiently place the small ships while falling back on Cell-based logic exclusively for the larger ships. 

This would drastically reduce the size of the pre-generated candidate dictionary, balancing the memory footprint while maintaining the tight \gls{lp} bounds for the most computationally demanding pieces.

\subsection{High-Performance Computing (HPC) Stress Testing}\label{high-performance-computing-testing}
Finally, future work should deploy the Ship model onto a \gls{hpc} cluster. Bypassing the \SI{16}{\giga\byte} unified memory limit of the current testing hardware would allow researchers to track the absolute space-complexity breaking point of the candidate generation algorithm on vast, non-standard grid dimensions.
