% =============================================================================
% CHAPTER 2: Background
% =============================================================================
% SPDX-FileCopyrightText: 2026 Frank Langbein <frank@langbein.org>
% SPDX-License-Identifier: LPPL-1.3c
%
% Suggested sections: Context (or Wider context / Domain and context); Related
% work (or Literature review / Prior work); Problem statement (and research
% question(s)). Optional: Theory and concepts; Stakeholders and constraints
% (as subsections of Context or separate). End with a clear problem statement.
% Assume a competent reader; avoid lengthy descriptions of standard tools.
% =============================================================================

% This chapter provides the background needed to understand the project: the wider context, existing solutions or related work, and a clear problem statement (and research question(s) where applicable). Replace with your own review.

\section{Context}\label{sec:context}
% Wider context, domain, motivation. Optional subsections: stakeholders,
% constraints, assumptions.

This project is focussing on the game known as Battleship Solitaire. There is a common misconception that this game is the same as the 2-player probabilistic game popularised as a board game, it is not.

Battleship Solitaire is a deterministic logic puzzle, that players solve using the given hints and tallies provided on the rows and columns. It is a pure application of Discrete Tomography, which is the mathematical study of constructing a geometric object from its projections.

Solving discrete tomography isn't just for puzzles, it has real-world applications, mainly in medical imaging (such as MRI/CT scans), scheduling, and VLSI microchip design \citep{herman1999discrete}.

From a computational complexity perspective, Battleship Solitaire is formally classified as an NP-Complete problem \citep{sevenster2004battleships}, a class of problems first formalised by Karp~\citep{karp1972reducibility}. This classification indicates that as the grid dimensions ($N \times N$) scale, the state space of valid fleet configurations undergoes combinatorial explosion. For NP-Complete problems, no known algorithm can guarantee a solution in polynomial time. Instead, all deterministic algorithms currently available exhibit exponential time complexity ($O(2^n)$) in the worst-case scenario.

A guaranteed polynomial-time algorithm for Battleship Solitaire would only be possible if it were mathematically proven that $P = NP$, which remains one of the most profound open problems in theoretical computer science. Because worst-case exponential scaling is unavoidable under current theoretical constraints, computational research instead must focus on the average-case performance.

This establishes the critical need for advanced algorithmic techniques, such as \gls{ilp} and branch-and-bound search~\citep{wolsey1998integer}. These methods do not alter the worst-case exponential boundary, but they intelligently prune the massive search space to find feasible solutions in practical, real-world timeframes.

\section{Related work}\label{sec:related}
% Prior solutions, methods, or systems; identify gaps your work addresses.
% Optional subsections: by theme, chronology, or type (e.g.\@ systems, methods, theory).

Several algorithmic approaches exist for solving NP-Complete problems, with Constraint Programming (CP) being one and Integer Linear Programming (ILP) being another. 

CP uses domain reduction to cross out infeasible values for each square in a puzzle. \gls{ilp} translates the entire board into a matrix, then uses branch-and-bound algorithms~\citep{wolsey1998integer} to mathematically find the optimal layout.

The foundation literature for this project is the \gls{ilp} formulations proposed by Meuffels \citep{meuffels-battleship}. He demonstrated that it is possible to conceptualise two distinct ways to translate the Battleship Solitaire puzzle into ILP mathematical models. The paper defines the cell-based and ship-based approaches, but lacks extensive testing and scaling of the problems they solve. This project aims to turn these models into a solution written in the modern day, with testing completed that pushes the boundaries of combinatorial algorithms to understand them further.

\citet{sevenster2004battleships} provides the foundational complexity-theoretic treatment, proving that the puzzle is NP-Complete.

This project uses the Gurobi Optimizer \citep{gurobi2024}, which is the industry standard for solving optimisation problems. 

The project also uses \gls{csplib} \citep{csplib:prob014}, which contains many constraint problem datasets. Allowing the solver to read the format provided by \gls{csplib} means the research is reproducible.

\section{Problem statement}\label{sec:problem}
% Clear statement of the problem and, if applicable, research question(s).

The aim of this project is to solve the NP-Complete Battleship Solitaire, using \gls{ilp} paradigms, to determine the most computationally efficient ILP approach.