% =============================================================================
% Appendix: Supplementary Material
% =============================================================================
% SPDX-FileCopyrightText: 2026 Frank Langbein <frank@langbein.org>
% SPDX-License-Identifier: LPPL-1.3c
%
% Appendices hold supporting material (detailed tables, derivations, survey text,
% user-study materials) so the main narrative stays focused. Complete source
% code is usually submitted separately, not in the report. Use \chapter and
% \section as in main chapters. Replace with your own supplementary material.
% =============================================================================

This appendix contains the supporting material that would interrupt the flow of the main chapters: the \gls{csplib} parser regular expression with worked examples, the derivation of the Big-M constants used in the linearised degree constraints of Section~\ref{subsec:degree-constraints}, and the full per-class output of the candidate generator.

\section[CSPLib Prolog parser]{\gls{csplib} Prolog parser}\label{app:parser}

The parser used in Section~\ref{subsec:data-structures-parsing} consumes the standard \gls{csplib} Prolog format published with Problem 014. Each instance is a single \verb|problem(...)| term with five comma-separated fields: an instance identifier, a fleet specification, a row tally vector, a column tally vector, and a list of hints with one-indexed coordinates.

A representative instance from \verb|data/csplib.pl|:
\begin{verbatim}
    problem(113,
        [4:1,3:2,2:3,1:4],
        [2,4,3,3,2,4,1,1,0,0],
        [0,5,0,2,2,3,1,3,2,2],
        [c@[7,10],w@[1,6]]).
\end{verbatim}

The five fields are extracted by a single compiled regular expression. Whitespace is tolerated around all commas, brackets, and parentheses; the inner lists may be empty (in which case the puzzle has no hints):
\begin{verbatim}
    pattern = re.compile(
        r'problem\s*\(\s*(\d+)\s*,'      # instance id
        r'\s*\[(.*?)\]\s*,'              # fleet
        r'\s*\[(.*?)\]\s*,'              # row tallies
        r'\s*\[(.*?)\]\s*,'              # column tallies
        r'\s*\[(.*?)\]\s*\)\.',          # hints
        re.DOTALL
    )
\end{verbatim}

Hints are then extracted from the fifth capture group with a secondary regular expression \verb|([wcmlrtb])@\[\s*(\d+)\s*,\s*(\d+)\s*\]|. \gls{csplib} uses one-indexed coordinates throughout, which the parser converts to zero-indexed coordinates before constructing the \verb|BattleshipPuzzle| object.

\section{Big-M derivation for the linearised degree constraints}\label{app:bigm}

The Big-M constants $M = 4$ for the submarine bound, $M = 3$ for the end-piece upper bound, $M = 1$ for the end-piece lower bound, and $M = 2$ for the middle-piece bounds in Section~\ref{subsec:degree-constraints} are not arbitrary. Each is the smallest value that fully deactivates the corresponding constraint when its indicator is zero, given the bound $0 \leq \eta_{i,j} \leq 4$.

\textbf{Submarine Bound:}
The constraint $\eta_{i,j} \leq M (1 - v_{i,j}^c)$ must be a no-op when $v_{i,j}^c = 0$, i.e. it must be implied by $\eta_{i,j} \leq |\mathcal{N}(i, j)| \leq 4$ alone. The smallest $M$ satisfying $M \geq \max \eta_{i,j} = 4$ is therefore $M = 4$.

\textbf{End-piece Upper Bound:}
The constraint $\eta_{i,j} \leq 1 + M(1 - e_{i,j})$ must be a no-op when $e_{i,j} = 0$. This requires $1 + M \geq 4$, giving $M \geq 3$, so the smallest valid value is $M = 3$.

\textbf{End-piece Lower Bound:}
The constraint $\eta_{i,j} \geq 1 - M(1 - e_{i,j})$ must be a no-op when $e_{i,j} = 0$. This requires $1 - M \leq 0$, giving $M \geq 1$, so the smallest valid value is $M = 1$.

\textbf{Middle-piece bounds.}
For both the upper bound $\eta_{i,j} \leq 2 + M(1 - v_{i,j}^m)$ and the lower bound $\eta_{i,j} \geq 2 - M(1 - v_{i,j}^m)$ to be no-ops when $v_{i,j}^m = 0$, we require $2 + M \geq 4$ and $2 - M \leq 0$, both giving $M \geq 2$. The smallest valid value is therefore $M = 2$ on both sides.

Choosing the smallest valid $M$ matters: a tighter Big-M produces a tighter linear-programming relaxation, which strengthens the bounds Gurobi computes at the root node and reduces the cuts it has to add downstream.

\section{Candidate generator output for the standard fleet}\label{app:candidate-counts}

The Ship Model's pre-processing routine of Section~\ref{subsec:pre-proc-and-candidate-gen} enumerates every geometrically valid ship placement on an $N \times N$ grid for each fleet class $L \in F$. For the standard \gls{csplib} fleet $F = \{4{:}1, 3{:}2, 2{:}3, 1{:}4\}$ the per-class counts on a $10 \times 10$ grid are:

\begin{table}[h]
  \centering
  \begin{tabular}{lccc}
    \toprule
    Length $L$ & Horizontal placements & Vertical placements & Total \\
    \midrule
    4 & 70 & 70  & 140 \\
    3 & 80 & 80  & 160 \\
    2 & 90 & 90  & 180 \\
    1 (submarines) & 100 & --- & 100 \\
    \midrule
    \multicolumn{3}{r}{Grand total $|\mathcal{C}|$:} & 580 \\
    \bottomrule
  \end{tabular}
  \caption{Per-class candidate counts for the standard \gls{csplib} fleet on a $10 \times 10$ grid. Submarines are recorded only once because their orientation is irrelevant. Total of 580 candidates matches the test assertion in \texttt{src/solver\_utils.py}.}
\end{table}

The generator's growth rate is $\mathcal{O}(N^2 L_{\max})$ across all fleet classes combined. On the $30 \times 30$ scaled fleet this rises to several thousand candidates, which is the primary source of the Ship Model's memory footprint discussed in Section~\ref{subsec:hardware-memory-ceilings}.

